Train Harder, Smarter:\\Using Graph Theory to Create Route Suggestions

### 2013-12-03 paper.tex
Serious updates to the paper.

\documentclass[twocolumn,11pt]{article}

@@

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[toc,page]{appendix}
\usepackage{url}

@@

@@

\section{Background}

@@

First and foremost, we opted to use the Open Street Map format, which is open
source, fairly robust, and has a very active community that contributes to both
the maps themselves and to various (mostly open-source) software projects that
utilize the format. More information on this format can be found on their
website, openstreetmap.org.

Additionally, we leveraged some existing code to more easily be able to create
a demonstrable algorithm. Pyroute was used, since it has the very nice feature
that the navigation component was entirely decoupled into PyrouteLib. This
allows easy utilization of the Pyroute tools for GUI setup and OSM download
while also making it rather explicit where the navigation functions are. The
OSM wiki pages have a good deal of information
on Pyroute and PyrouteLib; refer to them for more information.

@@

The first weight is based on distance. The idea is that there is a theoretical,
perfect loop that covers exactly the amount of distance that you would like,
and the algorithm does its best to direct you along that loop. See
Figure~\ref{fig:ideal} for an example of what this might look like.In order to
direct you along this loop, weights are generated based on
direction. We'll consider the method for finding the optimal direction,
$\theta$.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \draw (2,0) circle(2);
    \filldraw (0,0) circle(0.05);
    \node at (0.4,0.1) {$x_0$};

@@

    \filldraw (4,0) circle(0.05);
    \node at (4.2,0.2) {$x_j$};

    \node at (1.8,-0.7) {$x_i$};
    \node at (2.7,-1.4) {$x_{\mbox{{\tiny ideal}}}$};

@@

The next step is to utilize these equations to calculate $\theta$. At any point
along the algorithm, we are at some $x_i$, and we have travelled a distance,
$d_i$. Figure~\ref{fig:algvals} should help to understand what exactly the
algorithm is attempting to achieve. The idea is this: if we have travelled some
distance $d_i$, then there is an $x_{\mbox{{\tiny ideal}}}$ on the ``ideal''
loop. The arc between $x_{\mbox{{\tiny ideal}}}$ and $x_0$ will be of length
$d_i$. The algorithm searches for a point $x_j$ such that the arc between
$x_{\mbox{{\tiny ideal}}}$ and $x_j$ will have the \emph{same} distance as the
line between $x_i$ and $x_j$. Fortunately, we have framed the equations for
calculating $x_j$ in terms of the distance travelled along the arc, and the
distance between $x_i$ and $x_j$ can easily be calculated using the pythagorean
theorem. Note that we find $x_j$ as follows:
\begin{align}
  x_j = \frac{d}{2\pi} ( & \cos( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \cos \phi ) + x_0
                                  \label{eq:xj} \\
  y_j = \frac{d}{2\pi} ( & \sin( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \sin \phi ) + y_0.
                                  \label{eq:yj}
\end{align}
To satisfy our constraint, we have
\begin{equation}
d_j = \sqrt{ ( x_j(d_j) - x_i )^2 + ( x_j(d_j) - x_i )^2 }. \label{eq:dj}
\end{equation}

\subsubsection{Finding $d_j$}
Obviously, this is a rather difficult equation to truly solve, so we must only
heuristically find a solution. Unfortunately, even the heuristic solution is
a difficult one for a few reasons. Most problematically, the sine and cosine
functions have up to two stationary points on the interval we are considering
and there may not be a solution of $d_j$ that will satisfy the equality.
Nonetheless, we will do our best to mitigate these issues and apply a good
heuristic to find $d_j$. First, we rearrange our equation so that we can apply
a root-finding approximation algorithm to it:
\[ f(d_j) = \sqrt{(x_j(d_j) - x_i)^2 + (y_j(d_j) - y_i)^2} - d_j. \]
This simply allows the $d_j$ we are looking for to be the zero of the function.
The first exciting bit of news is that for most solutions of this equation with
$d_j < d$, it will work out that $f(0) > 0$ and $f(d-d_i) < 0$.
We know that $f(0) \geq 0$ by the
positivity of norms, and the square root of the sum of the squares is the
euclidian norm. We also can be certain that $f(d-d_i) < 0$ since
the norm is also unable to change by more than the diameter
of the ideal loop, or even less based on the proximity of $x_i$ to the center of
the loop. In situations where $f(d-d_i) > 0$, we notice that the distance to the
start of the loop is further away than the amount of distance we have left to
travel; this means that we should just route back to the start anyways. As such,
we have bounds inside of which we know that there exists a solution. We apply
Brent's method, a robust root approximation algorithm, to determine the solution
inside of this zone. The idea behind Brent's method is to combine the bisection
method, the secant method, and inverse quadratic interpolation. Readers
unfamiliar with any of these techniques should refer to CITATIONS. Some
pseudocode is given for Brent's Method in Algorithm~\ref{alg:brent}. The idea
is this:

@@

The algorithm details out the actual swapping of $a_k$ and $b_k$, but here
we'll just assume that $|f(a_k)| > |f(b_k)|$ at all points in time; that is,
$b_k$ is a better approximation of the solution than $a_k$. We also require
$b_{k-1}$; for now, we'll set it to $a_k$.

@@

\begin{algorithm}
  \caption{Using Brent's Method to find a zero of a function}
  \label{alg:brent}
  \begin{algorithmic}
    \Require $f(a_0)f(b_0) > 0$
    \Function{Brent}{$f : [a,b] \to \mathbb{R}$, $a_0$, $b_0$}
      \If{$|f(a_0)| < |f(b_0)|$}
        \Call{swap}{$a_0$,$b_0$}
      \EndIf
      \State $b_{k-1} \gets a_0$
      \State $B \gets$ True \Comment{bisected last time?}
      \While{$f(b_k) \not = 0$ {\bf and} $|b_k - a_k| > \epsilon$}
      \If{$|f(a_k)| < |f(b_k)|$}
      \Call{swap}{$a_k$,$b_k$}
      \EndIf
      \State $b_{k+1} \gets$\Call{inv\_quad\_interp}{$f,a_k,b_k,b_{k-1}$}
      \State $B$ $\gets$ \Call{need\_bisect}{$a_k$,$b_{k-2}$,$b_{k-1}$,$b_k$,$b_{k+1}$,$B$}
      \If{$B$}
        \State $\displaystyle b_{k+1} \gets \frac{a_k+b_k}{2}$ \Comment{bisection method}
      \EndIf
      \State $b_{k-2} \gets b_{k-1}$
      \State $b_{k-1} \gets b_k$
      \If{$f(a_k)f(s) < 0$}~~$b_k \gets s$
      \Else~~$a_k \gets s$
      \EndIf
      \EndWhile
    \EndFunction
  \end{algorithmic}
  \begin{algorithmic}
    \Function{need\_bisect}{$a_k$,$b_{k-2}$,$b_{k-1}$,$b_k$,$b_{k+1}$,$B$}
      \If{$\displaystyle b_{k+1} \not \in \left[ \frac{3a_k+b_k}{4},b_k \right]$}
      \State \Return True
      \ElsIf{$B$ is True}
      \State \Return
        $\displaystyle |b_{k+1}-b_k| \geq \frac{|b_k-b_{k-1}|}{2}$
      \State ~~~~~~~~~~~$\land$~~$|b_k-b_{k-1}| < |\delta|$
      \Else~
      \Return $\displaystyle |b_{k+1}-b_k| \geq \frac{|b_{k-1}-b_{k-2}|}{2}$
      \State ~~~~~~~~~~~~~~$\land$~~$|b_{k-1}-b_{k-2}| < |\delta|$
      \EndIf
    \EndFunction
  \end{algorithmic}
  \begin{algorithmic}
    \Function{inv\_quad\_interp}{$f$,$a$,$b$,$c$}
      \If{$f(a) \not= f(c)$ {\bf and} $f(b) \not= f(c)$}
        \State \Return $\displaystyle \frac{af(b)f(c)}{(f(a)-f(b))(f(a)-f(c))}$
        \State ~~~~~~~~~~~$\displaystyle + \frac{bf(a)f(c)}{(f(b)-f(a))(f(b)-f(c))}$
        \State ~~~~~~~~~~~$\displaystyle + \frac{cf(a)f(b)}{(f(c)-f(a))(f(c)-f(b))}$
      \Else
        \State \Return $\displaystyle b - f(b)\frac{b-a}{f(b)-f(a)}$
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

There are a few more important characteristics to consider when calculating
weights. Perhaps the easiest of these to deal with are the built in road types
that the Open Street Map format provides. It is easy enough to assign these
priorities based on what a cyclist would reasonably hope to see. Basically,
busy roads get weighted lower than low-traffic roads, but things like
pedestrian walkways and other unridable roads also get high weights.

Another characteristic to consider is whether a road has already been
travelled. Repeatedly riding along the same stretch of road can make a ride
rather boring, so by weighting against this, it can be avoided.

\subsection{Using the Weights Together}

Now that we have developed all of the weight calculations, we can consider using
them to generate a route. Algorithm~\ref{alg:grow} lays out the basic idea in
pseudocode. It omits definitions for many of the functions because they were
covered in sufficient detail in previous sections. The general procedure is
to loop until the algorithm returns to the start point after at least 90\%
of the distance, $d$ is travelled. At each iteration, $\theta$ will be
calculated using \eqref{eq:theta}, where $(x_j,y_j)$ are calculated by applying
Algorithm~\ref{alg:brent} on \eqref{eq:dj}. Once $\theta$ is calculated, a
weight is generated for each node adjacent to $x_i$, and one of the nodes is
selected at random based on their weights.

@@

      \State $w \gets 0$ \Comment{sum of all weights}
      \For{$x \in$ \Call{edges}{$x_k$}}
        \State $w \gets w +$\Call{weight}{$x$}
      \EndFor
      \State $q \gets$ \Call{random}{$[0,w]$} \Comment{to choose node}
      \For{$x \in$ \Call{edges}{$x_k$}}
        \If{$q < $ \Call{weight}{$x$,$\theta$}}
          \State $x_{k+1} \gets x$
          \State $d_{k+1} \gets d_k +$\Call{dist}{$x_k$,$x_{k+1}$}
          \State $X \gets X \cup \{x_{k+1}\}$
          \State break
        \Else~$q \gets q -$\Call{weight}{$x$,$\theta$}
        \EndIf
      \EndFor
    \EndWhile
    \State \Return $X$
  \EndFunction

@@

\section{Moving to Three Dimensions} \label{sec:latlong}

As mentioned earlier, using the Euclidian norm for latitudes and longitudes
is highly inaccurate; latitudes and longitudes are spherical coordinates with
fixed $r$, so considering the latitude and longitude to be $x$ and $y$ values is
a very flawed idea. As one can see in Figure~\ref{fig:sphere}, the contrast is
stark. Latitude and longitudes are angles, not positions in space.

\begin{figure}
  \includegraphics[width=3in]{images/558px-3D_Spherical.png}
  \caption{The Spherical Coordinate System} \label{fig:sphere}
\end{figure}

\subsection{Distance Calculations}
The haversine formula is used to calculate distances between two points on the
surface of a sphere. The haversine formula uses the haversin function,
which is defined as follows:
\newcommand{\haversin}{\mbox{haversin}}
\[ \haversin (\theta) = \sin^2(\frac{\theta}{2}). \]
Given two coordinate pairs, $(\phi_1, \lambda_1)$ and $(\phi_2, \lambda_2)$,
and a radius, $r$, the haversine formula calculates the distance $d$ between
the two coordinate pairs:
\begin{align*}
  d = r\haversin^{-1}(&\haversin(\phi_2 - \phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2 -\lambda_1)
\end{align*}
Using the definition of the haversin function, this can be reduced into a
function in terms of elementary trigonometric functions. The only non-trivial
thing to note is that $\haversin^{-1}(x) = 2\sin^{-1}(\sqrt{x})$. Armed with
this knowledge, it is fairly easy to reduce. We'll use an intermediate variable
to hold the value under the square root:
\begin{align*}
  h = & \haversin(\phi_2-\phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2-\lambda_1) \\
  = & \sin^2\left(\frac{\phi_2 - \phi_1}{2}\right) + \\
  & \cos(\phi_1)\cos(\phi_2)\sin^2\left(\frac{\lambda_2-\lambda_1}{2}\right)\\
  d = & r\haversin^{-1}\left(\sqrt{h}\right) \\
  = & 2r\sin^{-1}\left(\sqrt{h}\right)
\end{align*}
By simply substituting this as the right hand side of \eqref{eq:dj}, many of the
projection issues can be avoided.

@@

However, the actual ideal loop and, consequently,
\eqref{eq:xj} and \eqref{eq:yj} depend on a set of parametric equations along
a cartesian coordinate plane. These are an inaccurate representation of the
intended ideal loop; as the loop progresses further from the equator, it will
become increasingly squashed in the east/west direction.

\begin{figure}
  \centering
  \includegraphics[width=2in]{images/220px-Numerical_aperture.png}
  \caption{The aperture of a cone} \label{fig:aperture}
\end{figure}
To find the proper loop, a cone must be projected from the origin of the sphere,
and the intersection of the cone and the sphere can be used to calculate the
ideal loop. The following equation describes a cone,
\[ 0 = (u \cdot d)^2 - |d||u|\cos \phi , \]
where $u= (x,y,z)$, $d = (a,b,c)$ is a vector parallel to the axis of the
cone, and $\phi$ is one-half the aperture of the cone. In
Figure~\ref{fig:aperture}, the $\theta$ pictured is the same as the $\phi$ in
this equation. Additionally, recall
the equation for a sphere:
\[ r = |u|. \]
For the most part, we can consider $r = 1$, since we are dealing with the
surface of a sphere. However, we do have to make certain calculations using
the actual radius of the earth. For example, in order to calculate the
aperture $\phi$, we realize that the length of the sides of the cone will be the
radius of the earth. Then $\sin \phi = \frac{r_c}{r}$, so we can calculate
$\phi$ as follows:
\begin{equation}
  \phi = \sin^{-1} \left( \frac{\frac{d}{2\pi}}{r} \right) \label{eq:phi}
\end{equation}

Unfortunately, the equations for spheres and cones are of little aid, since
they require
input in the form of cartesian coordinates. We would like to convert these to
a set of equations in spherical form. However, these equations become rather
complicated; for now, we've accepted the projected circle as ``good enough.''

@@

\section{Discussion}

>>>>

\documentclass[twocolumn,12pt]{article}

@@

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[toc,page]{appendix}

@@

\newcommand{\mycite}[1]{[\citenum{#1}]}

@@

\section{Background} \label{sec:back}

@@

First and foremost, we opted to use the Open Street Map format, which is open
source, fairly robust, and has a very active community that contributes to both
the maps themselves and to various (mostly open-source) software projects that
utilize the format. The interested reader should refer to \citenum{wik:osm} for
almost all information regarding OSM, including how to contribute, how to use,
and links to projects that utilize OSM instead of proprietary mapping software.

Additionally, we leveraged some existing code to more easily be able to create
a demonstrable algorithm. Pyroute was used, since it has the very nice feature
that the navigation component was entirely decoupled into PyrouteLib. This
allows easy utilization of the Pyroute tools for GUI setup and OSM download
while also making it rather explicit where the navigation functions are.
Notably, pyroute has generally been abandoned in favor of the more advanced
Open Source Routing Machine (OSRM) project, but pyroute was chosen for this
paper instead due to its relative ease of prototyping and intuitiveness. Readers
interested in OSRM should refer to \mycite{wik:osrm}; those interested in
pyroute can sate their interest with \mycite{wik:pyroute}. One notes that
pyroute uses the relatively simple $A^*$ algorithm for routing between two
points. This is a slower implementation than OSRM, but our usage of the actual
routing algorithm is minimal; for the most part, we use pyroute only for loading
maps and displaying routes on them.

@@

The first weight is based on distance. The idea is that there is a theoretical,
perfect loop that covers exactly the amount of distance that you would like,
and the algorithm does its best to direct you along that loop. See
Figure~\ref{fig:ideal} for an example of what this might look like. In order to
direct you along this loop, weights are generated based on
direction. We'll consider the method for finding the optimal direction,
$\theta$.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \draw (2,0) circle(2);
    \filldraw (0,0) circle(0.05);
    \node at (0.4,0.1) {$\bm{x_0}$};

@@

    \filldraw (4,0) circle(0.05);
    \node at (4.2,0.2) {$\bm{x_j}$};

    \node at (1.8,-0.7) {$\bm{x_i}$};
    \node at (2.7,-1.4) {$\bm{x}_{\mbox{{\tiny ideal}}}$};

@@

The next step is to utilize these equations to calculate $\theta$. At any point
along the algorithm, we are at some $\bm{x_i}$, and we have travelled a
distance, $d_i$. Figure~\ref{fig:algvals} should help to understand what exactly
the algorithm is attempting to achieve. The idea is this: if we have travelled
some distance $d_i$, then there is an $\bm{x}_{\mbox{{\tiny ideal}}}$ on then
``ideal'' loop. The arc between $\bm{x}_{\mbox{{\tiny ideal}}}$ and $\bm{x_0}$
will be of length $d_i$. The algorithm searches for a point $x_j$ such that the
arc between $\bm{x}_{\mbox{{\tiny ideal}}}$ and $\bm{x_j}$ will have the
\emph{same} distance as the line between $\bm{x_i}$ and $\bm{x_j}$. Fortunately,
we have framed the equations for calculating $\bm{x_j}$ in terms of the distance
travelled along the arc, and the distance between $\bm{x_i}$ and $\bm{x_j}$ can
easily be calculated using the pythagorean theorem. Note that we find $\bm{x_j}$
as follows:
\begin{align}
  x_j = \frac{d}{2\pi} ( & \cos( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \cos \phi ) + x_0
                                  \label{eq:xj} \\
  y_j = \frac{d}{2\pi} ( & \sin( \frac{ 2 \pi }{ d } ( \min( d, d_i + d_j ) )
                                  + \phi ) \notag \\ & - \sin \phi ) + y_0.
                                  \label{eq:yj}
\end{align}
To satisfy our constraint, we have
\begin{equation}
d_j = \sqrt{ ( x_j(d_j) - x_i )^2 + ( x_j(d_j) - x_i )^2 }. \label{eq:dj}
\end{equation}

\subsubsection{Finding $d_j$}
Obviously, this is a rather difficult equation to truly solve, so we must only
heuristically find a solution. Unfortunately, even the heuristic solution is
a difficult one for a few reasons. Most problematically, the sine and cosine
functions have up to two stationary points on the interval we are considering
and there may not be a solution of $d_j$ that will satisfy the equality.
Nonetheless, we will do our best to mitigate these issues and apply a good
heuristic to find $d_j$. First, we rearrange our equation so that we can apply
a root-finding approximation algorithm to it:
\[ f(d_j) = \sqrt{(x_j(d_j) - x_i)^2 + (y_j(d_j) - y_i)^2} - d_j. \]
This simply allows the $d_j$ we are looking for to be the zero of the function.
The first exciting bit of news is that for most solutions of this equation with
$d_j < d$, it will work out that $f(0) > 0$ and $f(d-d_i) < 0$.
We know that $f(0) \geq 0$ by the
positivity of norms, and the square root of the sum of the squares is the
euclidian norm. We also can be certain that $f(d-d_i) < 0$ since
the norm is also unable to change by more than the diameter
of the ideal loop, or even less based on the proximity of $x_i$ to the center of
the loop. In situations where $f(d-d_i) > 0$, we notice that the distance to the
start of the loop is further away than the amount of distance we have left to
travel; this means that we should just route back to the start anyways. As such,
we have bounds inside of which we know that there exists a solution. We apply
Brent's method, a robust root approximation algorithm, to determine the solution
inside of this zone. The idea behind Brent's method is to combine the bisection
method, the secant method, and inverse quadratic interpolation. Readers
unfamiliar with any of these techniques should refer to \mycite{wik:sec},
\mycite{wik:bisect}, \mycite{wik:invquad}, \mycite{wik:brent} for a brief
overview of the techniques. A more interested reader might consider
\mycite{sauanal} for an in-depth analysis of the techniques as well as some
context for using and understanding them. Some pseudocode is given for Brent's
Method in Algorithm~\ref{alg:brent}, with some helper functions in
Algorithm~\ref{alg:bhelp}. The idea
is this:

\begin{algorithm}[t]
  \caption{Using Brent's Method to find a zero of a function}
  \label{alg:brent}
  \begin{algorithmic}
    \Require $f(a_0)f(b_0) > 0$
    \Function{Brent}{$f : [a,b] \to \mathbb{R}$, $a_0$, $b_0$}
      \If{$|f(a_0)| < |f(b_0)|$}
        \State \Call{swap}{$a_0$,$b_0$}
      \EndIf
      \State $b_{k-1} \gets a_0$
      \State $B \gets$ True \Comment{bisected last time?}
      \While{$f(b_k) \not = 0$ {\bf and} $|b_k - a_k| > \epsilon$}
      \If{$|f(a_k)| < |f(b_k)|$}
        \State \Call{swap}{$a_k$,$b_k$}
      \EndIf
      \State $b_{k+1} \gets$\Call{inv\_quad\_interp}{$f,a_k,b_k,b_{k-1}$}
      \State $B$ $\gets$ \Call{nds\_bi}{$a_k$,$b_{k-2}$,$b_{k-1}$,$b_k$,$b_{k+1}$,$B$}
      \If{$B$} \Comment{use bisection method}
        \State $\displaystyle b_{k+1} \gets \frac{a_k+b_k}{2}$
      \EndIf
      \State $b_{k-2} \gets b_{k-1}$
      \State $b_{k-1} \gets b_k$
      \If{$f(a_k)f(s) < 0$}
        \State $b_k \gets s$
      \Else
        \State $a_k \gets s$
      \EndIf
      \EndWhile
    \EndFunction
  \end{algorithmic}
\end{algorithm}
\begin{algorithm}[t!]
  \caption{Helper functions for Brent's method} \label{alg:bhelp}
  \begin{algorithmic}
    \State \Comment{\emph{Determines whether to use bisection~~~~~~~}}
    \State \Comment{\emph{method or inverse quadratic~~~~~~~~~~~~~~~~~}}
    \State \Comment{\emph{interpolation~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~}}
    \Function{nds\_bi}{$a_k$,$b_{k-2}$,$b_{k-1}$,$b_k$,$b_{k+1}$,$B$}
      \If{$\displaystyle b_{k+1} \not \in \left[ \frac{3a_k+b_k}{4},b_k \right]$}
      \State \Return True
      \ElsIf{$B$ is True}
      \State \Return
        $\displaystyle |b_{k+1}-b_k| \geq \frac{|b_k-b_{k-1}|}{2}$
      \State ~~~~~~~~~~~$\land$~~$|b_k-b_{k-1}| < |\delta|$
      \Else
      \State \Return $\displaystyle |b_{k+1}-b_k| \geq \frac{|b_{k-1}-b_{k-2}|}{2}$
      \State ~~~~~~~~~~~$\land$~~$|b_{k-1}-b_{k-2}| < |\delta|$
      \EndIf
    \EndFunction
  \end{algorithmic}
  \begin{algorithmic}
    \State \Comment{\emph{returns the next iterate under inverse~~~~~}}
    \State \Comment{\emph{quadratic interpolation~~~~~~~~~~~~~~~~~~~~~~~~~}}
    \Function{inv\_quad\_interp}{$f$,$a$,$b$,$c$}
      \If{$f(a) \not= f(c)$ {\bf and} $f(b) \not= f(c)$}
        \State \Return $\displaystyle \frac{af(b)f(c)}{(f(a)-f(b))(f(a)-f(c))}$
        \State ~~~~~~~~~~~$\displaystyle + \frac{bf(a)f(c)}{(f(b)-f(a))(f(b)-f(c))}$
        \State ~~~~~~~~~~~$\displaystyle + \frac{cf(a)f(b)}{(f(c)-f(a))(f(c)-f(b))}$
      \Else
        \State \Return $\displaystyle b - f(b)\frac{b-a}{f(b)-f(a)}$
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The algorithm details out the actual swapping of $a_k$ and $b_k$, but here
we'll just assume that $|f(a_k)| > |f(b_k)|$ at all points in time; that is,
$b_k$ is a better approximation of the solution than $a_k$. We also require
$b_{k-1}$; at the beginning, $a_k$ will suffice instead.

@@

@@

There are a few more important characteristics to consider when calculating
weights. Perhaps the easiest of these to deal with are the built in road types
that the Open Street Map format provides. It is easy enough to assign these
priorities based on what a cyclist would reasonably hope to see. Basically,
busy roads get weighted lower than low-traffic roads, but things like
pedestrian walkways and other unridable roads also get high weights.
Interestingly, pyroute already sets weights for these values, but these are
generated based on preferences for navigating; these vary greatly from
preferences for going out and having a good time. A good example of this is the
weight that pyroute assigns for cycling on steps, which is not 0. It is
certainly possible for the cyclist to dismount and climb the steps, but one who
is riding for fun is highly unlikely to be excited about this prospect.

Another characteristic to consider is whether a road has already been
travelled. Repeatedly riding along the same stretch of road can make a ride
rather boring, so by weighting against this, it can be avoided. In order to deal
with this, the algorithm stores stretches of road that have been travelled and
halves the weight of any path that the rider has already ridden.

\subsection{Using the Weights to Generate a Route}

Now that we have developed all of the weight calculations, we can consider using
them to generate a route. Algorithm~\ref{alg:grow} lays out the basic idea in
pseudocode. It omits definitions for many of the functions because they were
covered in sufficient detail in previous sections. In particular the function
for calculating weights is omitted, but it utilizes the algorithms and
characteristics discussed in previous sections
The general procedure is
to loop until the algorithm returns to the start point after at least 90\%
of the distance, $d$ is travelled. At each iteration, $\theta$ will be
calculated using \eqref{eq:theta}, where $(x_j,y_j)$ are calculated by applying
Algorithm~\ref{alg:brent} on \eqref{eq:dj}. Once $\theta$ is calculated, a
weight is generated for each node adjacent to $x_i$, and one of the nodes is
selected at random based on their weights.

@@

      \State $w \gets 0$ \Comment{sum of all weights}
      \For{$x \in$ \Call{edges}{$x_k$}}
        \State $w \gets w +$\Call{weight}{$x$,$\theta$}
      \EndFor
      \State $q \gets$ \Call{random}{$[0,w]$}
      \For{$x \in$ \Call{edges}{$x_k$}}
        \If{$q < $ \Call{weight}{$x$,$\theta$}}
          \State $x_{k+1} \gets x$
          \State $d_{k+1} \gets d_k +$\Call{dist}{$x_k$,$x_{k+1}$}
          \State $X \gets X \cup \{x_{k+1}\}$
          \State break
        \Else~$q \gets q -$\Call{weight}{$x$,$\theta$}
        \EndIf
      \EndFor
    \EndWhile
    \State \Return $X$
  \EndFunction

@@

\section{Using Spherical Coordinates} \label{sec:latlong}

As mentioned earlier, using the Euclidian norm for latitudes and longitudes
is inaccurate; latitudes and longitudes are spherical coordinates  with
fixed $r$. As one can see in Figure~\ref{fig:sphere},
spherical coordinates are the three dimensional
equivalent of polar coordinates. Unfortunately, this poses two problems,
since we have only described generating a route around the origin in a two
dimensional plane. The first problem is switching to a spherical coordinate
system; the second is switching to three dimensions. Clearly,
considering the latitude and longitude to be values on a cartesian
coordinate plane is a flawed idea. Latitudes and longitudes are angles, not
positions in space.

\begin{figure}
  \centering
  \includegraphics[height=2in]{images/sphere.png}
  \caption{The Spherical Coordinate System~\mycite{img:spher}} \label{fig:sphere}
\end{figure}

\subsection{Distance Calculations}
Fortunately, distances are very well defined under any coordinate system, so we
can tackle both problems here without much ado. In the spherical coordinate
system, The haversine formula is used to calculate distances between two points
on the surface of a sphere. The haversine formula uses the haversin function,
which is defined as follows:
\newcommand{\haversin}{\mbox{haversin}}
\[ \haversin (\theta) = \sin^2(\frac{\theta}{2}). \]
Given two coordinate pairs, $(\phi_1, \lambda_1)$ and $(\phi_2, \lambda_2)$,
and a radius, $r$, the haversine formula calculates the distance $d$ between
the two coordinate pairs \mycite{wik:haversin}:
\begin{align*}
  d = r\haversin^{-1}(&\haversin(\phi_2 - \phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2 -\lambda_1)
\end{align*}
Using the definition of the haversin function, this can be reduced into a
function in terms of elementary trigonometric functions. The only non-trivial
thing to note is that $\haversin^{-1}(x) = 2\sin^{-1}(\sqrt{x})$. Armed with
this knowledge, it is fairly easy to reduce. We'll use an intermediate variable
to hold the value under the square root:
\begin{align*}
  h = & \haversin(\phi_2-\phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2-\lambda_1) \\
  = & \sin^2\left(\frac{\phi_2 - \phi_1}{2}\right) + \\
  & \cos(\phi_1)\cos(\phi_2)\sin^2\left(\frac{\lambda_2-\lambda_1}{2}\right)\\
  d = & r\haversin^{-1}\left(\sqrt{h}\right) \\
  = & 2r\sin^{-1}\left(\sqrt{h}\right)
\end{align*}
By simply substituting this as the right hand side of \eqref{eq:dj}, many of the
projection issues can be avoided.

 However, it still isn't perfect; the haversine
formula does not account for variations in the elevation of the earth. Instead,
it assumes one is travelling at sea level. Additionally, it assumes the earth is
a perfect sphere, which is also slightly inaccurate. Less accurate formulae
include flat-surface projection formulae, while more accurate formulae
generate geodesics along the surface of an ellipsoid that models the earth
\mycite{wik:dist}. For our purposes, however, the haversine formula will be
sufficiently accurate.

@@

Distance calculations are the easier half of the problem;
the actual ideal loop and, consequently,
\eqref{eq:xj} and \eqref{eq:yj} depend on a set of parametric equations along
a cartesian coordinate plane. We'll consider a few different methods for
converting to spherical coordinates.

\subsubsection{A Na\"{i}ve Approach} \label{sec:naiveloop}

The first, and perhaps most obvious, method is to simply use latitude and
longitude as the $x$ and $y$ in \eqref{eq:xj} and \eqref{eq:yj}. This is
not a superb solution, but it does have a few advantages, chief of which are
the relative ease of computation and development. For small loops relatively
close to the equator, it may even be fairly similar to the intended ideal loop.

However, the larger the loop and the further from the equator the more squashed
the loop will begin to be. One also notes that the conversion does require a
slight change to the original formulae: one cannot use $d$ as the radius of the
circle. Instead, one must use an approximate conversion of the input distance
to an angular delta. Latitudinal distances change very little (they only
change because the Earth is an ellipsoid, not a sphere), and, as such, the
nautical mile was derived as the distance between a minute of latitude
\mycite{wik:nautmil}. Thus, we can easily convert between a distance goal and an
approximate latitudinal radius. Unfortunately, this will be dramatically
incorrect the closer to the poles we get.

\subsubsection{Accurate Approach}

\begin{figure}
  \centering
  \includegraphics[width=2in]{images/220px-Numerical_aperture.png}
  \caption{The half-aperture of a cone \mycite{img:aperture}} \label{fig:aperture}
\end{figure}

To find the real ideal loop, a cone must be projected from the origin of the
sphere, and the intersection of the cone and the sphere can be used to calculate
the ideal loop. Note that this approach, while far superior to the flattened
approach, still does not compensate for the ellipsoidal nature of the Earth, or
for the variations in the elevation on the surface of the earth.

The following equation describes a cone:,
\[ 0 = (\bm{u} \cdot \bm{d})^2 - |\bm{d}|^2|\bm{u}|^2\cos^2 \phi , \]
where $\bm{u}= (x,y,z)$, $\bm{d} = (a,b,c)$ is a vector parallel to the axis of
the cone, and $\phi$ is one-half the aperture of the cone \mycite{wik:cones}. In
Figure~\ref{fig:aperture}, the $\theta$ pictured is the same as the $\phi$ in
this equation; we have used $\phi$ instead of the more accepted $\theta$ to
avoid confusion with the zenith in spherical coordinates.
Additionally, recall the equation for a sphere:
\[ r = |\bm{u}|. \]
For the most part, we can consider $r = 1$, since we are dealing with the
surface of a sphere. However, we do have to make certain calculations using
the actual radius of the earth. For example, in order to calculate the
aperture $\phi$, we realize that the length of the sides of the cone will be the
radius of the earth. Then $\sin \phi = \frac{r_c}{r}$, so we can calculate
$\phi$ as follows:
\begin{equation}
  \phi = \sin^{-1} \left( \frac{\frac{d}{2\pi}}{r} \right) \label{eq:phi}
\end{equation}

With $r = 1$, we do our best to create an equation for the circle of interest:
\begin{align*}
  0 = & (\bm{u}\cdot \bm{d})^2 - |\bm{d}|^2|\bm{u}|^2\cos^2 \phi \\
    = & (ax + by + cz)^2 - \cos^2 \phi \\
    = & a^2x^2 + b^2y^2 + c^2z^2 \\
    & + 2abxy + 2acxz
     + 2bcyz - \cos^2 \phi \\
\end{align*}

Unfortunately, these equations for spheres and cones are of little aid. The
 parametric equations for a circle are inspired by the polar coordinate form
of a circle, so we convert these equations to spherical form. Ideally, this will
likewise inspire a set of parametric equations for this circle.
\begin{align*}
  0 = & a^2x^2 + b^2y^2 + c^2z^2 \\
      & 2abxy + 2acxz + 2bcyz - \cos^2 \phi \\
    = & a^2\cos^2 \psi + b^2\sin^2 \psi + c^2\sin^2 \theta \\
      & + 2ab\cos \psi\sin \psi + 2ac\cos \psi \sin \theta \\
      & + 2bc\sin \psi \sin \theta  - \cos \phi
\end{align*}
Unfortunately, this doesn't make much headway. Without further inspiration,
we'll stick with the na\"{i}ve approach discussed in
Section~\ref{sec:naiveloop}.

@@

At first, the algorithm was rather unsuccessful. The generated weights may have
been useful, but a few things kept them from being very successful. First, the
weights were calculated at every node. This means that at every point along a
road, it would be determined if one should continue down the road or turn around
and head back in the other direction. Unfortunately, since many roads are rather
curvy, the weights could often tell the user to oscillate back and forth on a
single road for an entirely uninteresting ride.

In subsequent iterations of the algorithm, this was noticed and adjusted for;
instead, the next intersection is found, and the weights are calculated for the
point at the next intersection. This also allows for weights for previously
travelled sections of road to be compressed to pairs of nodes for each stretch
between an intersection.

Another issue arose with the preliminary calculation of $\theta$. At first, a
much more na\"{i}ve approach was taken. Instead of using Brent's method to find
the zeroes of \eqref{eq:dj}, the algorithm simply divided the search space into
one hundred sections and chose the section that was closest.
